A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-04 inst/manuscript/manuscript.tex
background edits

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a smapling frame according to
a sampling design. A sampling frame is the set of all units available to
be sampled. A sampling design assigns a positive probability of
inclusion (inclusion probability) to each unit in the sampling frame.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sampling, and cluster sampling. If a
sampling design selects units from the sampling frame while ignoring
their spatial locations, we call them ``Independent Random Sampling''
(IRS) designs. If a sampling design selects units from the sampling
frame while incorporating their spatial locations, we call them
spatially balanced designs. Spatially balanced designs can be obtained
using the Generalized Random Tessellation Stratified (GRTS) algorithm
(Stevens and Olsen, 2004), which we discuss in more detail in Section
\ref{subsec:spb_design}. The design-based approach combines the
randomness of the sampling design and the data collected via the sample
to estimate fixed, unknown parameters (e.g., means and totals) of a
population.

Treating the data as fixed and incorporating randomness through the
sampling design yields estimators having very few other assumptions.
Confidence intervals for these types of estimators are typically derived
using limiting arguments that incorporate all possible randomizations of
sampling units selected via the sampling design. Means and totals, for
example, are asymptotically normally distributed (normal) by the Central
Limit Theorem (under some assumptions). If we repeatedly sample the
surface, then 95\% of all 95\% confidence intervals constructed from a
procedure with appropriate coverage will contain the true, fixed mean.
Särndal et al. (2003) and Lohr (2009) provide thorough reviews of the
design-based approach.

The model-based approach assumes the data are a random realization of a
data-generating stochastic process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement random sampling.

Instead of estimating fixed but unknown parameters like a mean or total
(as in the design-based approach), the goal of model-based inference in
the spatial context is often to predict a realized variable, or value.
For example, suppose the realized mean of all population units is the
value of interest. Instead of \emph{estimating} a fixed, unknown mean,
we are \emph{predicting} the value of the mean, a random variable.
Prediction intervals are then derived using assumptions of the data
generating process. If we repeatedly generate the response values from
the same spatial process and sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data. A visual comparison of the design-based and
model-based assumptions is provided in Figure \ref{fig:fig1} (Ver Hoef
(2002) and Brus (2021) provide similar figures).

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples of size four. The response values at each site are fixed, but we obtain different estimates for the mean response because the randomly sampled sites vary from sample to sample. In the bottom row, we have three realizations of the same spatial process sampled at the same locations. The spatial process generating the response values has a single mean, but the realized mean is different in each of the three panels.}\label{fig:fig1}
\end{figure}

@@

The design-based approach can be used to select samples that are
``well-spread'' in space, or spatially balanced. Spatially balanced
samples are useful because parameter estimates from these samples tend
to vary less than parameter estimates from samples that are not
spatially balanced (Barabesi and Franceschi, 2011; Benedetti et al.,
2017; Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004). To
quantify the spatial balance of a sample, Stevens and Olsen (2004)
proposed loss metrics based on Voroni polygons. After the GRTS algorithm
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use the Generalized Random Tessellation Stratified
(GRTS) algorithm to select spatially balanced samples sampling because
the algorithm has several attractive properties. It accommodates finite
and infinite sampling frames, equal, unequal, and proportional (to size)
inclusion probabilities, legacy (historical) sampling (Foster et al.,
2017), a minimum distance between units in a sample, and replacement
units in a sample (which are units that can be sampled in place of an
original unit that can no longer be sampled). The GRTS algorithm samples
from finite and infinite populations by utilizing a mapping between
two-dimensional and one-dimensional space. The units in the
two-dimensional sampling frame are divided into cells using a
hierarchical address. This hierarchical address is then used to map the
units from two-dimensional space to a one-dimensional line where each
unit's line length equals its inclusion probability. A systematic sample
is conducted on the line and linked back to a unit in two-dimensional
space, which results in the desired sample. Stevens and Olsen (2004)
provides further details.

After selecting a spatially balanced sample using the GRTS algorithm
(i.e., a GRTS sample), data are collected and used to estimate
population parameters. To unbiasedly estimate population means and
totals from sample data, one can use the Horvitz-Thompson estimator
(Horvitz and Thompson, 1952). If \(\tau\) is a population total, the
Horvitz-Thompson estimate of \(\tau\), denoted by \(\hat{\tau}_{ht}\),
is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th unit in the sample
and \(\pi_i\) is the inclusion probability of the \(i\)th unit in the
sample. An estimate of the population mean can be obtained by dividing
\(\hat{\tau}_{ht}\) by the number of population units, \(N\).

While the Horvitz-Thompson estimator is unbiased for population means
and totals, it is also important to quantify the uncertainty in these
estimates. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but these estimators have two
drawbacks. First, they rely on calculating \(\pi_{ij}\), the probability
that unit \(i\) and unit \(j\) are both in the sample -- this quantity
can be challenging if not impossible to calculate analytically. Second,
these estimators ignore the spatial locations of the units in the
sampling frame. To address these two drawbacks simultaneously, Stevens
and Olsen (2003) proposed the local neighborhood variance estimator. The
local neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) compared to variance estimators ignoring spatial
locations, yielding narrower confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
process. Ver Hoef (2008) gives details on the theory of FPBK, but some
of the basic principles are summarized below. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to predict some linear function of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights. For
example, if we want to predict the population total across all
population units, then we would use a vector of 1's for the weights.

We often only have a sample of the \(N\) population units. Denoting
quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with subscript \emph{u}, let

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation} where \(\mathbf{X}_s\) and \(\mathbf{X}_u\) are the
design matrices for the sampled and unsampled population units,
respectively, and \(\bm{\beta}\) is the parameter vector of fixed
effects.

Let \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for the
sampled and unsampled population units, respectively. We assume
\(E(\bm{\delta}) = \mathbf{0}\) and that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th elment of the matrix
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is dependent random error variance
measuring coarse-scale (correlated) variability, \(\sigma^2_{2}\) is the
independent random error variance measuring fine-scale (independent)
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the correlation, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). Often
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are called the partial sill and
nugget, respectively. Any spatial covariance function could be used in
the place of the exponential, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

@@

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. Other methods,
such as k-nearest-neighbors (Fix and Hodges, 1989; Ver Hoef and
Temesgen, 2013), random forest (Breiman, 2001), Bayesian models
(Chan-Golston et al., 2020), among others, could also be used to obtain
predictions for a mean or total from spatially correlated responses of a
finite population. We choose to use FPBK because it is faster than a
Bayesian approach and it was developed with theoretically-based variance
estimators of means and totals for spatial data, whereas random forests
and k-nearest-neighbors use ad-hoc variance estimators in most cases
(Ver Hoef and Temesgen, 2013); additionally, FBPK outperformed the other
methods in most scenarios.

>>>>

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame. A sampling
frame is the set of all units available to be sampled. Units from the
sampling frame are selected as part of the sample according to a
sampling design, which assigns a positive probability of inclusion
(inclusion probability) to each unit from the sampling frame. Some
examples of commonly used sampling designs include simple random
sampling, stratified random sampling, and cluster sampling. When
sampling designs incorporate spatial locations into sampling, we call
the resulting samples ``spatially balanced.'' One approach to selecting
spatially balanced samples is the Generalized Random Tessellation
Stratified (GRTS) algorithm (Stevens and Olsen, 2004), which we discuss
in more detail in Section \ref{subsec:spb_design}. When sampling designs
do not incorporate spatial locations into sampling, we call the
resulting samples ``non-spatially balanced.''

Fundamentally, the design-based approach combines the randomness of the
sampling design with the data collected via the sample to justify the
estimation and uncertainty quantification of fixed, unknown parameters
of a population (e.g., a population mean). Treating the data as fixed
and incorporating randomness through the sampling design yields
estimators having very few other assumptions. Confidence intervals for
these types of estimators are typically derived using limiting arguments
that incorporate all possible samples. Sample means, for example, are
asymptotically normal (Gaussian) by the Central Limit Theorem (under
some assumptions). If we repeatedly select samples from the population,
then 95\% of all 95\% confidence intervals constructed from a procedure
with appropriate coverage will contain the true, fixed mean. Särndal et
al. (2003) and Lohr (2009) provide thorough reviews of the design-based
approach.

The model-based approach assumes the data are a random realization of a
data-generating stochastic process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement some form of random sampling.

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of
\emph{estimating} a fixed, unknown mean, we are \emph{predicting} the
value of the mean, a random variable. Prediction intervals are then
derived using assumptions of the data-generating stochastic process. If
we repeatedly generate response values from the same data-generating
stochastic process and select samples, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide thorough reviews of model-based
approaches for spatial data. In Figure \ref{fig:fig1}, we provide a
visual comparison of the design-based and model-based approaches (Ver
Hoef (2002) and Brus (2021) provide similar figures).

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/fig1-1} \caption{A visual comparison of the design-based and model-based approaches. In the top row, there is one fixed population with nine population units and three random samples of size four (points circled are those sampled). The response values at each site are fixed, but we obtain different estimates for the mean response in each random sample. In the bottom row, there are three realizations of the same data-generating stochastic process that are all sampled at the same four locations. The ata-generating stochastic process has a single mean, but the mean of the nine population units is different in each of the three realizations.}\label{fig:fig1}
\end{figure}

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples (samples that incorporate spatial
locations of the population units and are ``well-spread'' is space).
Spatially balanced samples are useful because parameter estimates from
these samples tend to vary less than parameter estimates from samples
that are not spatially balanced (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens and Olsen, 2004; Wang et al., 2013). The first spatially
balanced sampling algorithm seeing widespread use is the Generalized
Random Tessellation Stratified (GRTS) algorithm (Stevens and Olsen,
2004). To quantify the spatial balance of a sample, Stevens and Olsen
(2004) proposed loss metrics based on Voronoi polygons (Direchlet
Tessellations). After the GRTS algorithm was developed, several other
spatially balanced sampling algorithms emerged, such as the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
Sampling (Benedetti and Piersimoni, 2017), and Halton Iterative
Partitioning Sampling (Robertson et al., 2018). In this manuscript, we
select spatially balanced samples using the Generalized Random
Tessellation Stratified (GRTS) algorithm because it has several
attractive properties. More specifically, the GRTS algorithm
accommodates finite and infinite sampling frames, equal, unequal, and
proportional (to size) inclusion probabilities, legacy (historical)
sampling (Foster et al., 2017), a minimum distance between units in a
sample, and replacement units (replacement units are population units
that can be sampled when a population unit originally selected can no
longer be sampled). The GRTS algorithm selects samples by utilizing a
particular mapping between two-dimensional and one-dimensional space
that preserves proximity relationships. Via this mapping, units in
two-dimensional space are partitioned using a hierarchical address. This
hierarchical address is used to map population units to a
one-dimensional line. On the one dimensional line, each population
unit's line length equals its inclusion probability. Then, a systematic
sample of population units is selected on the line, yielding desired
sample. Stevens and Olsen (2004) provides more technical details.

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimate of \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th population unit in
the sample and \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample. An estimate of the population mean is
obtained by dividing \(\hat{\tau}_{ht}\) by \(N\), the number of
population units.

It is also important to quantify uncertainty \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and populatoin unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically.
Second, these estimators ignore the spatial locations of the population
units. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) and yield narrower confidence intervals compared to
variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next (for more technical details, see Ver Hoef (2008)) Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal one). Denoting quantities that are part of the sampled population
units with a subscript \emph{s} and quantities that are part of the
unsampled population units with subscript \emph{u}, let

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation} where \(\mathbf{X}_s\) and \(\mathbf{X}_u\) are the
design matrices for the sampled and unsampled population units,
respectively, \(\bm{\beta}\) is the parameter vector of fixed effects,
and \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for the
sampled and unsampled population units, respectively.

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean zero and a spatial correlation structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative (between zero and one), second-order
stationary (depending only on the distance between population units),
isotropic (independent of direction), and decay with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\), though Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or both. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (for a
thorough list of spatial covariance functions, see Cressie (1993). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013), random forests (Breiman, 2001), Bayesian
models (Chan-Golston et al., 2020), among others, could also be used to
obtain predictions for a mean or total from spatially correlated
responses of a finite population. Compared to the k-nearest-neighbors
and random forest approach, we prefer FBPK because it is model-based and
relies on theoretically-based variance estimators leveraging the model's
spatial covariance structure, whereas k-nearest-neighbors and random
forests use ad-hoc variance estimators (Ver Hoef and Temesgen, 2013).
Additionally, Ver Hoef and Temesgen (2013) studied compared FBPK,
k-nearest-neighbors, and random forests in a variety of spatial data
contexts, and FBPK tended to perform best. Compared to the Bayesian
approach, we prefer FPBK mostly because it is much more computationally
efficient.
